Caring 15-minute neighbourhoods

### 2024-10-10 paper/Manuscript.qmd
Edits before submission

Though the term 15-Minute City is relatively new, neighbourhood planning practice aiming to foster community and local travel to amenities is not. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. Along different dimensions, these planning forms have been extensively critiqued [@talenSocialSciencePlanned2017]. In this respect, neighbourhood planning approaches such as the 15-Minute City concept are being re-embraced by some, but from bottom-up perspective using co-creation tools and meaningful resident participation to achieve equitable and justice neighbourhood forms [@mahmoudCocreationPathwayUrban2021]. How to equitably prescribe urban form that 'enables' contact with social opportunities (instead of 'engendering it' from the top-down) is still under debate. Questions remain, including: what destinations really matter?, what mode and travel time threshold is most appropriate?, and who benefits? For instance, these dimensions are under discussion among proximity-based planners who use accessibility-based tools [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022; @guzmanProximityEnoughCritical2024].

From another perspective, feminist geographers, urban planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. Gendered social norms and cultural factors play a role in how, where, with whom, what for, and how far we move. Women have higher activities related to care [@ilo_care_2018], no matter at what life stage they are [@garciaromanGenderDifferencesTime2022]. For all these reasons the spatial organization of activities has a different impact according to gender, the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-minute city and mobility of care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, Mobility of Care is shown to be a significant proportion of daily travel, characterizing the trip purpose of approximately 30% of adults’ trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips take on tremendous gendered and socio-economic importance, as women, and especially those from lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity more broadly-, mobility of care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

The objectives of this research are two-fold. First, we aim to theoretically bridge Mobility of Care with the 15-Minute City concept to re-imagine what local amenities matter and to define an associated accessibility and diversity of accessibility measures. Second, we take a data-driven approach to classify areas into 'caring 15-minute neighbourhoods' typologies and examine associated residential profiles in an empirical examination of the city of Hamilton, Canada.

@@

In the last decades, the need for more sustainable, healthier, and livable cities have received significant attention. Planners and decision-makers have proposed alternatives to mobility-based neighbourhood planning principles to ones that center proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. The 15-Minute City was introduced in @moreno_introducing_2021 along with four dimensions: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, and digitalisation. However, the framework presented has been criticized within academic and planning circles [@guzmanProximityEnoughCritical2024]. Critiques highlight how the concept falls short in addressing the pre-existing structural forces and individual characteristics that drive inequalities, influencing who benefits or could benefit the most from such a paradigm [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is an aspirational ideal, but how will existing mobility and accessibility inequalities be addressed. Without directed and context-specific solutions, this popular concept risks becoming an empty city branding exercise. [@pozoukidou15MinuteCityDecomposing2021; @gowerPlanningInnovationCity2022].

Reflecting the flexible and aspirational presentation of the 15-Minute City; cities that have adopted the concept have done so using a diverse range of definitions and tools along with indistinct universal approaches. A trail blazer was Portland, US in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. Adopted before the '15-Minute City' concept of today, it aims to foster an inclusive city development based on prosperity, education, health, and equity over a 30 year horizon. The plan aims to promote neighbourhood self-sufficiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017], similarly taking on an "all populations" approach. Later, other cities adopted proximity-based goals using similar neutral approaches. For example, the 15-Minute City plan announced during the Paris mayor Anne Hidalgo’s 2020 re-election campaign, which helped popularize the concept in today's public discourse, emphasized six key social functions that should be easily accessible from any location [@ville_de_paris_paris_2022]. These locations included: housing, work, health care, groceries, education and leisure. The 15-Minute City concept inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. @teixeiraClassifying15minuteCities2024's global review finds that the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap. In other words: the 15-Minute City is aspirational but how do we get there and who will benefit?

The past can shed light on cautionary tales. While the 15-Minute City concept is new, neighbourhood planning to improve society outcomes is not: in fact the literature has drawn parallels from the 15-Minute City to Clarence A. Perry's Neighbourhood Unit Concept (NUC) from the 1930s [@kissfazekasCircleParadigms15minute2022]. The NUC is a socio-spatial normative scheme widely adopted by government agencies (and the real-estate community) in the Western world after the Second World War [@talenSocialSciencePlanned2017; @solow1969concept]. Pairing well with the objectives of planning agencies at the time, Perry's NUC would allow for efficient mass-building of cellular units that prioritized the perceived functional needs of women and children: each unit providing proximate access to an elementary school and supporting community facilities, shopping, parks and housing [@talenSocialSciencePlanned2017; @brodyNeighbourhoodUnitConcept2013]. The NUC primarily prioritized local service provision, though Perry had confidence in good design's contribution to 'neighbourhood spirit' [@hall2014cities]. From the NUC's conception until the end of the 1960s, planners' aspirational attempts to prescribe social meaning to the neighbourhood's physical form had been criticized to near extinction [@talenSocialSciencePlanned2017]. By social scientists, the overestimation on the impact of built environment on social life was critiqued. By other planners, the inability to agree on the specificity of the neighbourhood (i.e., population size and the type, quality, and quantity of amenity) as well as how neighbourhood units are connected to other neighbourhoods and the rest of the region.  In response to these criticisms, neighbourhood planning proponents redefined deterministic terminology; their prescriptive physical form will 'enable' social contact with opportunities rather than 'engendering' it.

The redefined bottom-up approach to community and neighbourhood planning has been adopted by New Urbanists, from which the 15-Minute City stems [@kissfazekasCircleParadigms15minute2022]; however, the effectiveness of these bottom-up processes remain to be seen due to the contemporary nature of the concept. Though, related research can be examined. In recent years, the question of _how can physical form be planned to enable an improved quality of life, for whom, and with what outcomes_ has occupied the urban planning research agenda. For instance, the examination of low-income households access to transportation and their likelihood of gaining employment [@blumenbergDriveWorkRelationship2017; @bastiaanssenDoesBetterJob2022] or the relationship between children's access to public transit and participation in after-school activities [@palmRolePublicTransit2020]. A new wave of researchers and practitioners focused on local and context-specific relationships with the proximity to destinations have also emerged [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022]. As reviewed in the city plans that have adopted 15-Minute City approaches, the NUC's criticisms, and recent urban planning research, the question of _how to enable improved quality of life through urban built environment_ has shown to be highly context-sensitive, with the need for further investigation.

@@

In examining how to enable improved quality of life through urban built environment, accessibility measures have become an increasingly popular tool. These measures are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. The 15-Minute City is an amenity-provision neighbourhood planning concept aimed at enabling the creation of urban environments that enhance life quality, making it well-suited for analysis using accessibility methodologies [@guzmanProximityEnoughCritical2024]. Recent works have applied accessibility measures to investigate the 15-Minute City across different geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied in many 15-Minute City examinations. This measure quantifies how many destinations can be reached from a point in space within a given travel time threshold, pairing well with examinations that have a normative travel time threshold, like the 15-Minute City. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and popularity among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, accessibility measures other than the cumulative opportunity have also been applied, reflecting the diversity of measures in the literature [@guzmanProximityEnoughCritical2024].

Another method that compliments the assessment of the 15-Minute City regarding the diversity of opportunity types is entropy. The entropy measure, based on the Shannon-Wiener index of species diversity, expresses relative evenness within a sequence [@whittaker1972evolution]. In urban studies and planning literature, the flexible concept of entropy has been used to characterize land-use mix diversity [@frankLinkingObjectivelyMeasured2005; @ewingTravelBuiltEnvironment2010a], particularly in the context of non-work trips [@cerveroTravelDemand3Ds1997], walking as a mode [@luUrbanDensityDiversity2017; @mavoaIdentifyingAppropriateLanduse2018], and suburban sprawl [@randallGISBasedLand2015] as relevant to this work. It has also been used to understand mobility behaviour [@mcbride2020exploration; @montero2023applying; @montero2023role]. However, work incorporating entropy indices alongside accessibility analysis are not exceptionally common. There are examples of their use as parameters within accessibility scores for employment opportunities [@chengMeasuringUrbanJob2013; @daiIncorporatingJobDiversity2018] and, more recently, to describe the diversity in transit facilities [@yinIncorporatingFacilityDiversity2024]. Within the conceptualisation of the 15-Minute City, diversity of land-use is a key feature. In this way, the entropy measure theoretically complements the 15-Minute City.

Classification algorithms also show promise as a tool in the assessment of the 15-Minute City, as they have been useful in identifying mobility and spatial typologies within the transportation planning literature. Often, these algorithms take the form of machine learning approaches: for instance, @burkeGeospatialAnalysisFramework2022 uses spatially constrained multivariate clustering, to develop urban form typologies related to the 15-Minute City. The k-means algorithm is another popular method: @ganUnderstandingUrbanMobility2020 applies it to analyse travel behaviour and classify metro stations based on mobility patterns. Similarly, [@montero2023role; @montero2023applying] cluster travel behavior profile typologies representing different population segments and place of residence in the Barcelona Metropolitan Area. Combing classification techniques is also common: for instance, @delmelleDifferentiatingPathwaysNeighborhood2017 uses self-organizing maps (SOM) algorithm to group U.S. neighbourhoods by minimizing dissimilarity in attributes before applying k-means for further classification. Similarly, @victorianoTimeSpaceMoney2020 uses SOM to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the Decision Tree algorithm to partition the data into interpretable classifications.

@@

Rather than focusing on _all_ destinations, it may valuable to examine those related to caring activities. This re-framing offers a feminist perspective on which urban functions matter, connecting well with the 15-Minute City concept. Caring activities, which fulfill the physical, psychological, and emotional needs of others, are among the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Conventionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care worldwide, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While an undercurrent of research had examined these unique household-serving trips over the decades, her seminal work was the first to consider all these trips as one category and demonstrate how mobility of care is a significant proportion of daily travel.

@@

Pairing well with the focus on shorter trips and the potential use of sustainable modes within the 15-Minute City, Mobility of Care trips are more local, shorter-distance and trip-chained. Compared to the trip to work, mobility of care trips are more frequently completed by foot, and less frequently by transit or bicycle [@ravensbergen2023exploratory]. However, little work to date examines walking to care destinations through the Mobility of Care framework. Though there is ample literature that examines walking to care destinations, such as schools (e.g., [@RN7;@RN8;@RN9]) and grocery stores (e.g., [@RN10;@RN11]), they tend to consider singular care destinations in research focusing on walkability.

In summary, we reviewed the neighborhood planning framework from which the 15-Minute City concept stems and is situated within. We then reviewed tools common and useful in investigating the 15-Minute City. Finally, we introduced the Mobility of Care concept, theoretically linking it to the 15-Minute City to highlight the importance of care destinations for all but especially along gendered and socio-economic lines. The objectives of this work are two-fold, achieved through an empirical examination of Hamilton, Canada, a mid-sized city:

>>>>

The term 15-Minute City is relatively new but the concept is not, as it is simply a normative cumulative-opportunities accessibility measure [@paez_measuring_2012]. Its newfound fame under the moniker of chronourbanism comes to complement long-standing efforts to foster community and local travel to amenities in neighbourhood planning practice. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. Along different dimensions, these planning forms have been extensively critiqued [@talenSocialSciencePlanned2017]. In this respect, neighbourhood planning approaches such as the 15-Minute City concept offer an opportunity for bottom-up approaches that leverage co-creation tools and meaningful resident participation to achieve equitable and just neighbourhood forms [@mahmoudCocreationPathwayUrban2021]. How to prescribe equitable urban forms that 'enable' contact with social opportunities (instead of 'engendering it' from the top-down) is still a matter for debate. Questions remain, including: what destinations really matter? What mode and travel time threshold is most appropriate? And who benefits? To be certain, these dimensions are under discussion among proximity-based planners who use accessibility-based tools [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022; @guzmanProximityEnoughCritical2024].

From another perspective, feminist geographers, urban planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990; @neto_gender_2015]. The causes for this are varied, and related to gendered social norms and cultural factors that play a role in how, where, with whom, what for, and how far we move. Of note is that women tend to travel more frequently and for shorter distance [@roorda_triap_2010; @morency_distance_2011] and have more activities related to care [@ilo_care_2018] no matter at what life stage they are [@garciaromanGenderDifferencesTime2022]. Combined, these findings indicate that the spatial organization of activities has a different impact according to gender, and therefore the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-Minute City and mobility of care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, care-related mobility can be shown to comprise a significant proportion of daily travel, characterizing approximately 30% of adults’ daily trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue that Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips are of particular gendered and socio-economic import, as women, and especially those in lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity more broadly-, mobility of care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

The objectives of this research are two-fold. First, we aim to theoretically bridge Mobility of Care with the 15-Minute City concept to re-imagine what local amenities matter from the perspective of care and to define an associated accessibility and diversity of accessibility measures. And secondly, we take a data-driven approach to classify areas into 'caring 15-minute neighbourhoods' typologies and examine associated residential profiles in an empirical examination of the city of Hamilton, Canada.

@@

In the last decades, the need for more sustainable, healthier, and livable cities has become a prominent concern. Planners and decision-makers have proposed a shift to neighbourhood planning principles centered on proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. The 15-Minute City was introduced in @moreno_introducing_2021 along with four dimensions: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, and digitalisation [q.v. @cerveroTravelDemand3Ds1997]. However, the framework presented has been criticized within academic and planning circles [e.g., @guzmanProximityEnoughCritical2024; @mouratidisTimeChallenge15minute2024] among other things for shortfalls in terms of addressing pre-existing structural forces and individual characteristics that drive inequalities and influence who benefits or could benefit the most from such an approach [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is an aspirational goal, but does not fully confront existing mobility and accessibility inequalities in need of redress. Without directed and context-specific solutions, this popular concept risks becoming an empty city branding exercise. [@pozoukidou15MinuteCityDecomposing2021; @gowerPlanningInnovationCity2022].

Reflective of the flexible and aspirational presentation of the 15-Minute City, the concept has been adopted by cities using a diverse range of definitions and tools along with indistinct universal approaches. A trail blazer in this respect was Portland (US) in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. Adopted before the '15-Minute City' concept of today, this plan aimed to foster inclusive city development based on prosperity, education, health, and equity over a 30 year horizon. Central to the plan was the promotion of neighbourhood self-sufficiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017], similarly taking on an "all populations" approach. Subsequently, other cities adopted proximity-based goals using similarly neutral approaches. For example, the 15-Minute City plan announced during the re-election campaign of Paris mayor Anne Hidalgo in 2020 emphasized six key social functions that should be easily accessible from any location [@ville_de_paris_paris_2022]. These locations included: housing, work, health care, groceries, education and leisure. The 15-Minute City concept inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. @teixeiraClassifying15minuteCities2024's global review finds that the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap. In other words: the 15-Minute City is aspirational but how do we get there and who will benefit?

The past can provide cautionary tales. While the 15-Minute City brand of planning is new, neighbourhood planning to improve society outcomes is not: in fact the literature has drawn parallels from the 15-Minute City to Clarence A. Perry's Neighbourhood Unit Concept (NUC) from the 1930s [@kissfazekasCircleParadigms15minute2022]. The NUC is a socio-spatial normative scheme widely adopted by government agencies (and the real-estate community) in the Western world after the Second World War [@talenSocialSciencePlanned2017; @solow1969concept]. Pairing well with the objectives of planning agencies at the time, Perry's NUC would allow for efficient mass-building of cellular units that prioritized the perceived functional needs of women and children: each unit providing proximate access to an elementary school and supporting community facilities, shopping, parks and housing [@talenSocialSciencePlanned2017; @brodyNeighbourhoodUnitConcept2013]. The NUC primarily prioritized local service provision, though Perry had confidence in good design's contribution to 'neighbourhood spirit' [@hall2014cities]. By the end of the 1960s, planners' aspirational attempts to prescribe social meaning to the neighbourhood's physical form had been criticized to near extinction [@talenSocialSciencePlanned2017]. A critique by social scientists was an overestimation of the impact of built environment on social life. Planners, on the other hand, could not reach consensus on the specificity of the neighbourhood (i.e., population size and the type, quality, and quantity of amenity) or how neighbourhood units connect between them and the rest of the region.  In response to these criticisms, neighbourhood planning proponents redefined their deterministic terminology; their prescriptive physical form would 'enable' social contact with opportunities rather than 'engendering' it.

The redefined bottom-up approach to community and neighbourhood planning was adopted  in the 1980s by American New Urbanists [@trudeau_new_2013], from which the 15-Minute City eventually stemmed [@kissfazekasCircleParadigms15minute2022]. However, the bottom-up effectiveness of these ideas remains to be seen due to the contemporary nature of the concept. Though, related research can be examined. In recent years, the question of _how can physical form be planned to enable an improved quality of life, for whom, and with what outcomes_ has occupied the urban planning research agenda. For instance, the examination of low-income households access to transportation and their likelihood of gaining employment [@blumenbergDriveWorkRelationship2017; @bastiaanssenDoesBetterJob2022] or the relationship between children's access to public transit and participation in after-school activities [@palmRolePublicTransit2020]. A new wave of researchers and practitioners focused on local and context-specific relationships with the proximity to destinations have also emerged [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022]. As reviewed in the city plans that have adopted 15-Minute City approaches, the NUC's criticisms, and recent urban planning research, the question of _how to enable improved quality of life through urban built environment_ is highly context-sensitive, prompting the need for further investigation.

@@

In examining how to enable improved quality of life through urban built environment, accessibility measures have become an increasingly popular tool. These measures are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. The 15-Minute City is an amenity-provision neighbourhood planning concept aimed at enabling the creation of urban environments that enhance life quality, making it well-suited for analysis using accessibility methodologies [@guzmanProximityEnoughCritical2024]. Recent works have applied accessibility measures to investigate the 15-Minute City across different geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied in many 15-Minute City examinations. This measure quantifies how many destinations can be reached from a point in space within a given travel time threshold , pairing well with normative examinations [see @paez_measuring_2012] that use travel time thresholds like x-minute cities [@logan_xminute_2022. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and popularity among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, accessibility measures other than the cumulative opportunity have also been applied, reflecting the diversity of measures in the literature [@guzmanProximityEnoughCritical2024].

Another concept that complements the assessment of the 15-Minute City regarding the diversity of opportunity types is entropy. The entropy measure, based on the Shannon-Wiener index of species diversity, expresses relative evenness within a sequence [@whittaker1972evolution]. In urban studies and planning literature, the concept of entropy has been used to characterize land-use mix diversity [@frankLinkingObjectivelyMeasured2005; @ewingTravelBuiltEnvironment2010a; @moniruzzaman_accessibility_2012] including to understand mobility behaviour [@mcbride2020exploration; @montero2023applying; @montero2023role], in the context of non-work trips [@cerveroTravelDemand3Ds1997], walking [@luUrbanDensityDiversity2017; @mavoaIdentifyingAppropriateLanduse2018], and suburban sprawl [@randallGISBasedLand2015]. However, entropy indices are rare in accessibility analysis. There are examples of their use as parameters within accessibility scores for employment opportunities [@chengMeasuringUrbanJob2013; @daiIncorporatingJobDiversity2018] and, more recently, to describe the diversity in transit facilities [@yinIncorporatingFacilityDiversity2024]. Given how 15-Minute Cities aim to provide access to numerous amenities, diversity of land-use is a key feature. In this way, the entropy measure theoretically complements the 15-Minute City.

Classification algorithms also show promise as a tool in the assessment of the 15-Minute City, as they have been useful in identifying mobility and spatial typologies within the transportation planning literature. Often, these algorithms take the form of machine learning approaches: in transportation, the use of Self-Organizing Maps (SOM) was pioneered by Delmelle [-@delmelle_spatial_2012] and has been used to group U.S. neighbourhoods by minimizing dissimilarity in attributes [@delmelleDifferentiatingPathwaysNeighborhood2017] and to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the Decision Tree algorithm to partition the data into interpretable classifications [@victorianoTimeSpaceMoney2020]. Other approaches include the use of spatially constrained multivariate clustering, to develop urban form typologies related to the 15-Minute City [@burkeGeospatialAnalysisFramework2022] and the use of k-means to analyse travel behaviour and classify metro stations based on mobility patterns[ @ganUnderstandingUrbanMobility2020].

@@

Rather than focusing on _all_ destinations, it is valuable to examine those related to caring activities. This framing offers a feminist perspective on urban functions that matter from a care perspective, and connects well with the 15-Minute City concept. Caring activities, which fulfill the physical, psychological, and emotional needs of others, are among the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Conventionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care worldwide, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While an undercurrent of research had examined these unique household-serving trips over the decades, her seminal work was the first to consider all these trips as one category and demonstrate how mobility of care is a significant proportion of daily travel.

@@

Pairing well with the focus on shorter trips and the potential use of sustainable modes within the 15-Minute City, Mobility of Care trips are more local, shorter-distance and trip-chained. Compared to the trip to work, mobility of care trips are more frequently completed by foot, and less frequently by transit or bicycle [@ravensbergen2023exploratory]. However, little work to date examines walking to care destinations through the Mobility of Care framework. Though there is ample literature that examines walking to care destinations, such as schools (e.g., [@RN7;@RN8;@RN9]) and grocery stores (e.g., [@RN10;@RN11]), they tend to consider singular care destinations in research focusing on walkability. 
In summary, we reviewed the neighborhood planning framework from which the 15-Minute City concept stems and is situated within. We then reviewed tools common and useful in investigating the 15-Minute City. Finally, we introduced the Mobility of Care concept, theoretically linking it to the 15-Minute City to highlight the importance of care destinations for all but especially along gendered and socio-economic lines. The objectives of this work are two-fold, achieved through an empirical examination of Hamilton, Canada, a mid-sized city:
